Time Stretching of the GeV Emission of GRBs: Fermi LAT data vs geometrical model

### 2014-08-13 Paper/gammaRays.tex
Corrections to Total Energy
Corrections to Distribution of stretching factors

To compute the total energy, we need to multiply the radiation intensity by frequency, and integrate it over frequencies $\omega$, volume $\left(r,\theta,\phi\right)$, and observer positions $\left(\sigma, \xi\right)$. We assumed $\theta \ll 1$, and have taken into account the same relativistic effects as we did in the observed particle count computation:
\begin{align}
E &= \int_0^\infty \dif \omega \int_0^\infty \dif r \int_0^\infty r \dif \theta \int_0^{2\pi} r \, \theta \dif \phi \int_0^\infty \dif \sigma \int_0^{2\pi} \sin\sigma \dif \xi \nonumber\\
&\quad {} \times \eta\left( r,\theta,\omega \right) \underbrace{\frac{1}{\gamma\left(1-v\cos\sigma \right)}}_{\text{time dilation}} \underbrace{\frac{1}{\gamma^2\left(1-v\cos\sigma \right)^2}}_{\text{aberration}} \underbrace{\frac{\omega}{\gamma\left(1-v\cos\sigma \right)}}_{\text{relativistic shift}} \\
&= \frac{4\pi^2}{\gamma^4} \int_0^\infty \dif \omega \, \omega \int_0^\infty \dif r \, r^2 \int_0^\infty \dif \theta \, \theta \, \eta\left( r,\theta,\omega \right) \int_0^\infty \dif \sigma \frac{\sin\sigma}{\left(1-v\cos\sigma \right)^4} \nonumber
\end{align}
An integral over $\sigma$ is computable analytically:
\begin{equation*}
\int_0^\infty \dif \sigma \frac{\sin\sigma}{\left(1-v\cos\sigma \right)^4} = \frac{2\left(3+v^2\right)}{3\left(1-v^2\right)^3} = \frac{2}{3}\gamma^6 \left(4 - \frac{1}{\gamma^2}\right)
\end{equation*}
Substituting this result into the expression for $E$ we get:
\begin{equation}
E = \frac{32\pi^2}{3}\left(\gamma^2 - \frac{1}{4}\right) \int_0^\infty \dif \omega \, \omega \int_0^\infty \dif r \, r^2 \int_0^\infty \dif \theta \, \theta \, \eta\left( r,\theta,\omega \right)
\end{equation}
Now we can use the expression for $\eta$ to do the remaining integrals:
\begin{align*}
E = \frac{32\pi^2 \eta_0}{3}\left(\gamma^2 - \frac{1}{4}\right) \int_0^\infty \dif \omega \, \omega \left(\frac{\omega}{\omega_0}\right)^\alpha \int_0^\infty \dif r \, \frac{r^2}{1 + \left(\frac{r}{r_0}\right)^n} \int_0^\infty \dif \theta \, \theta \, \exp\left(-\left(\frac{\theta}{\theta_0}\right)^2\left(\frac{\omega}{\omega_0}\right)^{-2k}\right)
\end{align*}
The integrals over $r$ and $\theta$ can be done symbolically:
\begin{equation*}
\int_0^\infty \dif r \, \frac{r^2}{1 + \left(\frac{r}{r_0}\right)^n} = \frac{\pi}{n\sin\left(\frac{3\pi}{n}\right)}r_0^3
\end{equation*}
\begin{equation*}
\int_0^\infty \dif \theta \, \theta \, \exp\left(-\left(\frac{\theta}{\theta_0}\right)^2\left(\frac{\omega}{\omega_0}\right)^{-2k}\right) = \frac{1}{2}\theta_0^2 \left(\frac{\omega}{\omega_0}\right)^{2k}
\end{equation*}
Now we only have one integral left:
\begin{equation}
E = \frac{16\pi^3}{3 n\sin\left(\frac{3\pi}{n}\right)} \left(\gamma^2 - \frac{1}{4}\right) \, \eta_0 \, r_0^3 \, \theta_0^2 \int_0^\infty \dif \omega \, \omega \left(\frac{\omega}{\omega_0}\right)^{2k+\alpha}
\end{equation}

@@

The second test is to calculate the distribution of the stretching factors of the observable bursts, and to compare it with observations.

The computation of the exact and precise distribution is technically hard (because stretching factors neither change monotonously with redshift, nor with off-axis angle), and computationally intensive. So we will not compute the precise distribution.

@@

We still have one ingredient missing though -- the evolution of the bursts density. We assume that the density is roughly proportional to the stars density, and the stars density is roughly proportional to the matter density, which changes with $z$ as $\left(1+z\right)^3$. It is clear, however, that since no stars existed at very small redshifts, the burst density should decline there, so we add an exponential cutoff to it:
\begin{equation}
\rho = \rho_0 \left(1+z\right)^3 \exp\left(-\frac{z}{z_c}\right)
\end{equation}
Here $\rho_0$ is a normalization factor, and $z_c$ is a redshift scale, after which the density is cut off.

@@

	As with redshifts, we get a properly distributed $\chi$ by solving the equation:
	\begin{align}
	\Phi_\chi\left(\chi\right) = \left(\frac{\chi}{\chi_\text{max}}\right)^2 = y \\
	\chi = \chi_\text{max}\sqrt{y}
	\end{align}
	Here $y$ is an another random variable uniformly distributed in the range $0$ to $1$.
}
\item{Now we should check, if the burst in a selected position can be observed: $p_\infty\left(z,\chi;\omega_2,\omega_3\right) > p_\text{min}$. If it is, add $\kappa\left(z,\chi\right)$ to the sample. If not, repeat from the step $3$.}
\item{If the sample is not as big as we want yet, repeat from the step $3$.}
\end{enumerate}

>>>>

To compute the total energy, we need to multiply the radiation intensity by frequency, and integrate it over frequencies $\omega$, volume $\left(r,\theta,\phi\right)$, and observer positions $\left(\sigma, \xi\right)$. We assumed $\theta \ll 1$, and have taken into account the same relativistic effects as we did in the observed particle count computation:
\begin{align}
E &= \int_0^\infty \dif \omega \int_0^\infty \dif r \int_0^\infty r \dif \theta \int_0^{2\pi} r \, \theta \dif \phi \int_0^\infty \dif \sigma \int_0^{2\pi} \sin\sigma \dif \xi \nonumber\\
&\quad {} \times \eta\left( r,\theta,\omega \right) \underbrace{\frac{1}{\gamma\left(1-v\cos\sigma \right)}}_{\text{time dilation}} \underbrace{\frac{1}{\gamma^2\left(1-v\cos\sigma \right)^2}}_{\text{aberration}} \underbrace{\frac{\omega}{\gamma\left(1-v\cos\sigma \right)}}_{\text{relativistic shift}} \\
&= \frac{4\pi^2}{\gamma^4} \int_0^\infty \dif \omega \, \omega \int_0^\infty \dif r \, r^2 \int_0^\infty \dif \theta \, \theta \, \eta\left( r,\theta,\omega \right) \int_0^\infty \dif \sigma \frac{\sin\sigma}{\left(1-v\cos\sigma \right)^4} \nonumber
\end{align}
An integral over $\sigma$ is computable analytically:
\begin{equation*}
\int_0^\infty \dif \sigma \frac{\sin\sigma}{\left(1-v\cos\sigma \right)^4} = \frac{2\left(3+v^2\right)}{3\left(1-v^2\right)^3} = \frac{2}{3}\gamma^6 \left(4 - \frac{1}{\gamma^2}\right)
\end{equation*}
Substituting this result into the expression for $E$ we get:
\begin{equation}
E = \frac{32\pi^2}{3}\left(\gamma^2 - \frac{1}{4}\right) \int_0^\infty \dif \omega \, \omega \int_0^\infty \dif r \, r^2 \int_0^\infty \dif \theta \, \theta \, \eta\left( r,\theta,\omega \right)
\end{equation}
Now we can use the expression for $\eta$ to do the remaining integrals:
\begin{align*}
E = \frac{32\pi^2 \eta_0}{3}\left(\gamma^2 - \frac{1}{4}\right) \int_0^\infty \dif \omega \, \omega \left(\frac{\omega}{\omega_0}\right)^\alpha \int_0^\infty \dif r \, \frac{r^2}{1 + \left(\frac{r}{r_0}\right)^n} \int_0^\infty \dif \theta \, \theta \, \exp\left(-\left(\frac{\theta}{\theta_0}\right)^2\left(\frac{\omega}{\omega_0}\right)^{-2k}\right)
\end{align*}
The integrals over $r$ and $\theta$ can be computed symbolically:
\begin{equation*}
\int_0^\infty \dif r \, \frac{r^2}{1 + \left(\frac{r}{r_0}\right)^n} = \frac{\pi}{n\sin\left(\frac{3\pi}{n}\right)}r_0^3
\end{equation*}
\begin{equation*}
\int_0^\infty \dif \theta \, \theta \, \exp\left(-\left(\frac{\theta}{\theta_0}\right)^2\left(\frac{\omega}{\omega_0}\right)^{-2k}\right) = \frac{1}{2}\theta_0^2 \left(\frac{\omega}{\omega_0}\right)^{2k}
\end{equation*}
Now we have only one integral left:
\begin{equation}
E = \frac{16\pi^3}{3 n\sin\left(\frac{3\pi}{n}\right)} \left(\gamma^2 - \frac{1}{4}\right) \, \eta_0 \, r_0^3 \, \theta_0^2 \int_0^\infty \dif \omega \, \omega \left(\frac{\omega}{\omega_0}\right)^{2k+\alpha}
\end{equation}

@@

The second test is to calculate the distribution of stretching factors of observable bursts, and to compare it to observations.

The computation of the exact and precise distribution is technically hard (because stretching factors change monotonously neither with redshifts, nor with off-axis angles), and computationally intensive. So we will not compute the precise distribution.

@@

We still have one ingredient missing though -- the evolution of the bursts density. We assume that the bursts density is roughly proportional to the stars density, and the stars density is roughly proportional to the matter density, which changes with $z$ as $\left(1+z\right)^3$. It is clear, however, that since no stars existed at very small redshifts, the burst density should decline there, so we add an exponential cutoff to it:
\begin{equation}
\rho = \rho_0 \left(1+z\right)^3 \exp\left(-\frac{z}{z_c}\right)
\end{equation}
Here $\rho_0$ is a normalization factor, and $z_c$ is a redshift scale, after which the density is cut off.

@@

	As with redshifts, we get a properly distributed $\chi$ by solving the equation:
	\begin{align}
	\Phi_\chi\left(\chi\right) = \left(\frac{\chi}{\chi_\text{max}}\right)^2 = y \\
	\chi = \chi_\text{max}\sqrt{y}
	\end{align}
	Here $y$ is an another independent random variable uniformly distributed in the range $0$ to $1$.
}
\item{Now we should check, if the burst in a selected position can be observed: $p_\infty\left(z,\chi;\omega_2,\omega_3\right) > p_\text{min}$. If it is, add $\kappa\left(z,\chi\right)$ to the sample. If not, repeat from the step $3$.}
\item{If the sample is not as big as we want yet, repeat from the step $3$.}
\end{enumerate}
